\section{Future Work}

After conducting this research some questions remain.

\subsection{Is the unit $\eta$ rule the only rule which requires a type-directed equality?}

In our presented theory, the only judgmental equality rule which required the type-directed approach was the unit $\eta$ rule.
\citet{altenkirch2001} present an algorithm for deciding the $\eta$ rule for the coproduct type.
However, it's unclear if their algorithm would fit with the type-directed approach we present here.
\citet{gilbert2019} present a type theory with a universe of definitionally irrelevant propositions.
While similar to the unit $\eta$ rule, their $\eta$ rule for propositions requires not only the type of the two terms be known, but also the type of the type.

\subsection{Could our implementation be optimized?}

While our smalltt implementation as is failed to perform as well as the syntax-directed smalltt, it's not out of the question that it could be optimized further.
We leave this optimization as future work.

\subsection{Was our method of correcting for variation in performance due to memory layout effective?}

Previous work has shown that a program's memory layout can drastically affect performance, with even small things like the values of environment variables affecting this layout.
We attempted to correct for this in our statistical test, but it's unclear if we did this properly.
Properly developing statistical methods to solve this problem would be very useful when running any sort of performance comparison between programs.
