\section{Results and Conclusion}

\begin{figure}[!htb]
  \centering
  \begin{tabular}{|c | c|}
    \hline
    Benchmark & P-Value
    \\ \hline
    asymptotics & $1.5 \times 10^{-161}$
    \\ \hline
    conv\_eval & $9.9 \times 10^{-261}$
    \\ \hline
    stlc & $8.2 \times 10^{-309}$
    \\ \hline
    stlc5k & $3.6 \times 10^{-305}$
    \\ \hline
    stlc10k & DNF
    \\ \hline
    stlc100k & DNF
    \\ \hline
    stlc\_lessimpl & $4.1 \times 10^{-311}$
    \\ \hline
    stlc\_lessimpl5k & $0$
    \\ \hline
    stlc\_lessimpl10k & DNF
    \\ \hline
    stlc\_small & $0$
    \\ \hline
    stlc\_small10k & $2.6 \times 10^{-284}$
    \\ \hline
    stlc\_small5k & $1.8 \times 10^{-301}$
    \\ \hline
  \end{tabular}
  \caption{Benchmark Results}
  \label{fig:results}
\end{figure}

In \autoref{fig:results} we can see the results of the performance comparison between the original syntax-directed smalltt and our modified type-directed smalltt.
The p-value is the probability that we got the observed benchmark results under the assumption that our modified version smalltt isn't slower than the original.
Across the board this probability is very small, and so it seems likely that our modified version of smalltt is slower than the original.
That being said, a few cases jump out in this table.

In the stlc\_lessimpl5k and stlc\_small benchmarks, the p-value is 0.
We interpret this as a p-value so small that it can't be represented as a double-precision floating point number, since these are what we used to do the statistical test.

Our modified version of smalltt did not finish the stlc10k, stlc100k, and stlc\_lessimpl10k benchmarks due to running out of memory.
That our implementation failed to complete some of the benchmarks isn't entirely surprising.
smalltt includes equivalent benchmarks for some other common theorem provers such as Agda, Coq, and Idris2.
In the performance comparison given in the README for smalltt, both Agda and Idris2 also failed to finish on the stlc100k.

This, however, doesn't detract from the fact that the original smalltt is clearly faster than our implementation.
Therefore, we can reject our thesis statement and say that while the type-directed equality supports more rules than the syntax-directed equality, the syntax-directed equality performs better.


